---
title: "ncGTW User Manual"
package: ncGTW
author: "Chiung-Ting Wu"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: biomed-central.csl
output: 
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{ncGTW User Manual}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{ncGTW}
    %\VignetteKeywords{mass spectrometry, metabolomics, alignment}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse=TRUE,
    comment="#>"
)
```

# Introduction

Neighbor-wise Compound-specific Graphical Time Warping (ncGTW) is an alignment
algorithm that can align LC-MS profiles by leveraging expected retention time
(RT) drift structures and compound-specific warping functions. This algorithm is
improved from graphical time warping (GTW) [@gtw16], a popular dynamic time
warping (DTW) based alignment method [@dtw90]. Specifically, ncGTW uses
individualized warping functions for different compounds and assigns constraint
edges on warping functions of neighboring samples. That is, ncGTW avoids the
popular but not accurate assumption which assumes all the m/z bins in the same
sample share the same warping function. This assumption often fails when the
dataset contains hundreds of samples or the data acquisition time longer than a
week. Moreover, by considering the RT drifts structure, ncGTW can align RT more
accurately.

`ncGTW` is an R package developed as a plug-in of `xcms`, a popular LC-MS data
analysis R package [@xcms06; @xcms08; @xcms10]. Due to the same warping function
assumption or bad parameter settings, `xcms` may have some misaligned features,
and there is a function in `ncGTW` to identify such misalignments. After
identifying the misaligned features, the user can realign these features with
the alignment function in `ncGTW` to obtain a better alignment result for more
accurate analysis, such as peak-regrouping or peak-filling with `xcms`.

You can install the latest version of ncGTW from GitHub by
```{r, eval = FALSE}
devtools::install_github("ChiungTingWu/ncGTW")
```
or from Bioconductor by
```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ncGTW")
```

# Quick Start

To check there are misaligned features from `xcms` or not, one can input two
`xcms` grouping results with different values of RT window parameter (`xcms`
grouping parameter, `bw`) to the function `misalignDetect()`. One value of `bw`
should be the expected maximal RT drift, and another should be near to the RT
sampling resolution (the inverse of scan frequency). If there are some detected
misaligned features, the user can decide to adjust the paramters in `xcms` or
use `ncGTW` to realign them. Besides the `xcms` aligment results, the only
paramter with no default in `misalignDetect()` is `ppm`, which should be set as
same as `ppm` of the peak detection in `xcms`.

```{r, eval = FALSE}
excluGroups <- misalignDetect(xcmsLargeWin, xcmsSmallWin, ppm)
```

# Misaligned Feature Detection and Realignment

## RT Structure Incorporation

To demonstrate the workflow of ncGTW, an example dataset is included in the
package. The aquisition time of the dataset is more than two weeks, in which the
20 samples are selected from a large dataset for a quick demonstration.

```{r, message = FALSE}
library(xcms)
library(ncGTW)
filepath <- system.file("extdata", package = "ncGTW")
file <- list.files(filepath, pattern = "mzxml", full.names = TRUE)
# The paths of the 20 samples
```

To incorporate the RT structure, the order of the paths in `file` should be as
same as the sample acquisition order (run order). In the example dataset, the
index in each file name is the acquisition order, so we sort the paths according
to `tempInd`. When dealing with other dataset, the user should make sure the
order of the paths is as same the order of data acquisition.

```{r}
tempInd <- matrix(0, length(file), 1)
for (n in seq_along(file)){
    tempCha <- file[n]
    tempLen <- nchar(tempCha)
    tempInd[n] <- as.numeric(substr(tempCha, regexpr("example", tempCha) + 7, 
        tempLen - 6))
}
file <- file[sort.int(tempInd, index.return = TRUE)$ix]
# Sort the paths by data acquisition order to incorporate the RT structure
```

## XCMS Preprocessing

As a plug-in, the inputs of `ncGTW` are the alignment results from `xcms`, so
first we need to apply `xcms` on the dataset. The parameters should be decided
by the user when dealing with other datasets.

```{r, message = FALSE, warning = FALSE}
CPWmin <- 2 
CPWmax <- 25 
ppm <- 15 
xsnthresh <- 3 
LM <- FALSE
integrate <- 2 
RTerror <- 6 
MZerror <- 0.05
prefilter <- c(8, 1000)
# XCMS parameters
ds <- xcmsSet(file, method="centWave", peakwidth=c(CPWmin, CPWmax), ppm=ppm, 
    noise=xsnthresh, integrate=integrate, prefilter=prefilter)
gds <- group(ds, mzwid=MZerror, bw=RTerror)
agds <- retcor(gds, missing=5)
# XCMS peak detection and RT alignment
```

To detect the misaligned features, `ncGTW` needs two XCMS grouping results with
different values of `bw`. The larger one should be expected maximal RT drift,
and the smaller one should be the RT sampling resolution (the inverse of scan
frequency).

```{r, message = FALSE}
bwLarge <- RTerror
bwSmall <- 0.3
# Two different values of bw parameter
xcmsLargeWin <- group(agds, mzwid=MZerror, bw=bwLarge)
xcmsSmallWin <- group(agds, mzwid=MZerror, bw=bwSmall, minfrac=0)
# Two resolution of XCMS grouping results
```

## ncGTW Workflow

After XCMS preprocessing, `ncGTW` can be applied on the results. There are two
major steps in `ncGTW`, misaligned feature detection and misaligned feature
realignment.


### Misaligned Feature Detection

To detect the misaligned features, `misalignDetect()` needs two different XCMS
grouping results as inputs. This function tells which features in `xcmsLargeWin`
could be broken into several small features in `xcmsSmallWin`, and the detected
features should be misaligned features. `ppm` is one criteria to decide the
small features in `xcmsLargeWin` are from the same compounds or not, and should
be set as same as the one in XCMS peak detection.

```{r, message = FALSE}
excluGroups <- misalignDetect(xcmsLargeWin, xcmsSmallWin, ppm)
# Detect misaligned features
show(excluGroups)
```

There are two peak groups (features) are detected as shown in `excluGroups`. 
Before realigning them, the raw profile of each detected feature of each sample 
needs to load from the files. `loadProfile()` loads the needed information with 
file paths (`file`) and the detected features (`excluGroups`) as inputs.

```{r, message = FALSE}
ncGTWinputs <- loadProfile(file, excluGroups)
# Load raw profiles from the files
```

The user can also check the detected features are really misaligned or not by
viewing the extracted ion chromatogram. `plotGroup()` draws the extracted ion
chromatogram. `ncGTWinputs` is the loaded information from `loadProfile()`,
`xcmsLargeWin@rt$corrected` is the alignment by XCMS, and `ind` is just a
parameter for indexing the chromatograms. The user are free to set `ind`.

```{r, message = FALSE}
for (n in seq_along(ncGTWinputs))
    plotGroup(ncGTWinputs[[n]], slot(xcmsLargeWin, 'rt')$corrected, ind=n)
    # (Optional) Draw the detected misaligned features
```

From the two figures, it is clear that these two features are really misaligned.
The color of curves changes from green, purple, to red according to the sample
run order.

### Misaligned Feature Realignment

After the needed information is loaded to `ncGTWinputs`, we can start to realign
the detected features with `ncGTW`. The parameter `parSamp` is for parallel
computing, which decides how many samples would be aligned together each time.
In this example, there are 20 samples, and `parSamp` are set as 5. Thus, there
would be four sub-groups of samples, and there are five samples in each
sub-group. Also, `bpParam` is set as four workers to align the four sub-groups
simultaneously. After all sub-groups are aligned, `ncGTW` would integrate the
four alignment results together to generate the final realignment.
If the user do not need parallel computing, `parSamp` could be set as same as
the total sample number. However, if sample number is larger than 100, it is
strongly recommended to split the samples into several sub-groups.

```{r, message = FALSE}
ncGTWoutputs <- vector('list', length(ncGTWinputs))
# Prepare the output variable
ncGTWparam <- new("ncGTWparam")
# Initialize the parameters of ncGTW alignment with default
for (n in seq_along(ncGTWinputs))
    ncGTWoutputs[[n]] <- ncGTWalign(ncGTWinputs[[n]], xcmsLargeWin, parSamp=5,
        bpParam=SnowParam(workers=4), ncGTWparam)
# Perform ncGTW alignment
```

After realignment, we need to send the realignment result to `adjustRT()` to
generate new RT warping functions to replace `xcmsLargeWin@rt$corrected`, and
send them back to `xcms` for further analysis.

```{r, message = FALSE}
ncGTWres <- xcmsLargeWin
# Prepare a new xcmsSet to contain the realignment result
ncGTWRt <- vector('list', length(ncGTWinputs))
for (n in seq_along(ncGTWinputs)){
    adjustRes <- adjustRT(ncGTWres, ncGTWinputs[[n]], ncGTWoutputs[[n]], ppm)
    # Generate the new warping functions
    peaks(ncGTWres) <- ncGTWpeaks(adjustRes)
    # Relocate the peaks to the new RT points according to the realignment.  
    ncGTWRt[[n]] <- rtncGTW(adjustRes)
    # Temporary variable for new warping functions 
}
```

Again, the user can also check the realignment by viewing the extracted ion
chromatogram with `plotGroup()`.

```{r, message = FALSE}
for (n in seq_along(ncGTWinputs))
    plotGroup(ncGTWinputs[[n]], ncGTWRt[[n]], ind = n)
    # (Optional) Draw the realigned features
```

From the two figures, it is clear that the two misaligned features now are
realigned accurately, comparing to the XCMS alignment.

## Peak-filling with Realigned RT

One of the most obvious impact of the realignment is the quality of peak-filling
in `xcms`. Due to the more accurate warping functions, the peak-filling step has
a higher change to retrieve the missing peaks back. That is, the guessing of the
positions of the missing peaks becomes more accurate according to the new
warping functions. Here we demonstrate the differences of peak-filling of the
two misaligned features.

```{r, message = FALSE}
groups(ncGTWres) <- excluGroups[ , 2:9]
groupidx(ncGTWres) <- groupidx(xcmsLargeWin)[excluGroups[ , 1]]
# Only consider the misaligned features
rtCor <- vector('list', length(file))
for (n in seq_along(file)){
    rtCor[[n]] <- vector('list', dim(excluGroups)[1])
    for (m in seq_len(dim(groups(ncGTWres))[1]))
        rtCor[[n]][[m]] <- ncGTWRt[[m]][[n]]
}
slot(ncGTWres, 'rt')$corrected <- rtCor
# Replace the XCMS warping function to ncGTW warping function
XCMSres <- xcmsLargeWin
groups(XCMSres) <- excluGroups[ , 2:9]
groupidx(XCMSres) <- groupidx(xcmsLargeWin)[excluGroups[ , 1]]
# Consider only the misaligned features with XCMS warping function
```

After extracting the misaligned features and replacing the old warping 
functions, we can apply `fillPeaks` in `xcms` for peak-filling. Since  
`fillPeaks` accepts only one warping function for each sample, we need to 
replace the function `fillPeaksChromPar()` first.

```{r, message = FALSE}
assignInNamespace("fillPeaksChromPar", ncGTW:::fillPeaksChromPar, ns="xcms",
    envir=as.environment("package:xcms"))
# Replace fillPeaksChromPar() in XCMS
ncGTWresFilled <- fillPeaks(ncGTWres)
XCMSresFilled <- fillPeaks(XCMSres)
# Peak-filling with old and new warping functions
compCV(XCMSresFilled)
compCV(ncGTWresFilled)
# Compare the coefficient of variation
```

For the first misaligned feature, the coefficient of variation (CV) decreases
from 0.369 to 0.229, and for the second one, the CV decreases from 0.351 to
0.119. Thus, it is very clear that new warping functions improve the quality of
peak-filling significantly.

# References
